Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-03 documents/ms.tex
small edits

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Timothy Morton\altaffilmark{\princeton},
   Suzanne Aigrain\altaffilmark{\oxford},
   Daniel Foreman-Mackey\altaffilmark{\uw,\sagan}
   \& Vinesh Rajpaul\altaffilmark{\oxford}
}

\altaffiltext{\columbia}{Simons Fellow, Department of Astronomy, Columbia
University, NY, NY, RuthAngus@gmail.com}
\altaffiltext{\princeton}{Department of Astrophysical Sciences, Princeton
University,  Princeton, NJ}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
UK}
\altaffiltext{\sagan}{Sagan Fellow}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process (GP) with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test this method's effectiveness, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

The brightness of a spotted, rotating star often varies in a non-sinusoidal and
Quasi-Periodic (QP) manner, due to active regions on its surface which
rotate in and out of view.
Complicated surface spot patterns produce non-sinusoidal variations,
and the finite lifetimes of these active regions and
differential rotation on the stellar surface produce quasi-periodicity.
A strictly periodic sinusoid therefore does not well describe
rotationally modulated stellar light curves.
A physically realistic model of the stellar surface
would ideally perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery when conditioned on the data.
However, such physical models require many free parameters in order to
accurately represent a stellar surface, and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that stars may have hundreds of spots, the number
of free parameters in such a model quickly becomes unwieldy, especially to explore
its posterior probability distribution function (PDF).
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results; however, such
relatively inflexible models sacrifice precision.

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram,
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010},
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its precise shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

@@

GPs are commonly used in the machine learning community and increasingly
in other scientific fields such as biology, geophysics and cosmology.
More recently, GPs have been used in stellar and exoplanet fields to capture
stellar variability or instrumental systematics \citep[see \eg][]{Gibson2012,
Haywood2014, Dawson2014, Barclay2015, Haywood2015, Evans2015, Rajpaul2015,
Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset.
An in-depth introduction to Gaussian processes is provided by
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points by means of a
parametric kernel function that models the autocorrelation behaviour of the
signal and defines the form of the covariance matrix.
As a qualitatively demonstration, we present the time-series
in Figure \ref{fig:GP_example}:
the \kepler\ light curve of KIC \kepexample, a relatively active star that
rotates once every $\sim$ 4 days, with stochastic variability typical of
\kepler\ FGK stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated, and points more
widely separated are loosely correlated.
A GP models the way in which this correlation varies with the separation between
data points; that is, it models not the data directly but rather the {\it
covariance structure}.
This lends GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas modelling the time series itself
might require much more complexity.
Figure \ref{fig:GP_example} demonstrates how a GP model fits the light curve of
KIC \kepexample.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, adequately fits the
KIC \kepexample\ light curve:
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}

However, The SE kernel function does not well describe the covariance
in stellar light curves, nor is it {\it useful} for the problem of
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.  For this reason, we use the `Quasi-Periodic' kernel.
This kernel is used by \citet{Rasmussen2005} to capture the variable behaviour
of CO$_2$ emissions from the Mauna Loa volcano in Hawai'i.
It is defined as
\begin{equation}
k_{i,j} = A \exp \left[-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2\left(\frac{2\pi}{P}\right) \right].
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing
points separated by half a period to be weakly correlated.
We also use an extra parameter $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This can be interpreted to represent underestimation of observational
uncertainties~---~if the uncertainties reported on the data are too small, it
will be non-zero~---~or it can capture any remaining ``jitter'' or residuals
that are not captured by the effective GP model.
%In practice, this parameter should always be non-zero when performing
%GP regression, because the covariance matrix must be positive definite and
%slight deviations from positive definiteness can arise due to most
%solvers using approximate, rather than exact, matrix inversion.
%Including a small amount of extra variance in the model ensures enough
%flexibility for matrix inversion algorithms to not run into such numerical
%issues.
We use this QP kernel function to produce the GP model
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.

@@

This paper is laid out as follows.
In \textsection \ref{sec:simulations} of this paper we describe the simulated
light curves used to test the three rotation period recovery
methods (ACF, LS periodogram and GP).
The GP method is described in \textsection \ref{sec:method} and the ACF and LS
periodogram methods are described in \textsection \ref{sec:comparison}.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data,
and the results are discussed in \textsection \ref{sec:discussion}.

@@

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We used \naigrain\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \naigraincurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \naigrain\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in \citet{Aigrain2015}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
A histogram of the \naigrain\ solid-body rotation periods is shown in figure
\ref{fig:period_hist}.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real \kepler\ light curves with
no obvious astrophysical variability in order to preserve the noise properties
of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated light curve with a
period of \aigrainexampleperiod days.

The ranges and distributions of the physical stellar parameters used in the
simulated light curves are tabulated below, in table
\ref{tab:simulation_parameters}, and figure \ref{fig:period_hist} shows the
distribution of rotation periods in the \citet{Aigrain2015} sample.

@@

In order to recover the rotation period of a simulated light curve using
Gaussian processes, we sample the following posterior PDF:
\begin{equation}
p({\bm \theta}\,|\,y) \propto \mathcal L(y\,|\,{\bm \theta}) p({\bm \theta}),
\end{equation}
\label{eq:posterior}
where $y$ are the light curve data, $\bm \theta$ are the hyperparameters
of the kernel described in equation \ref{eq:QP}, $\mathcal L$ is the
GP likelihood function, and $p({\bm \theta})$ is the prior on the
hyperparameters.  The GP likelihood is similar to the simple Gaussian likelihood
function used for optimisation problems where the uncertainties are
Gaussian and uncorrelated. The latter can be written
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\frac{(y_n-\mu)^2}{\sigma_n^2}
    - \frac{N}{2}\ln(2\pi\sigma_n^2),
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainites on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \mathrm{constant},
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
	\mathbf{C} &=& \left (\begin{array}{cccc}
	\sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
	\sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
	\sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!  Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function,
${\bf K}$ replaces ${\bf C}$ in the above equation.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
In order to assess the performance of the GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD\ \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/25.png}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    light curve in figure \ref{fig:demo_lc}.
$\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the uncertainties reported on the data are too small, this
parameter will be non-zero.
    The green line in the period panel shows the injected period.
    This figure was made using \texttt{corner.py} \citep{Corner}.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

\section{Comparison with standard methods}
\label{sec:comparison}
\subsection{ACF}

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MRD of the pgram-recovered periods is \percentpgramMAD\ \% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
There is a large number of drastically overestimated periods recovered using
the periodogram.
Excess power at long periods can be generated by long-term trends that are not
related to rotation.
This problem becomes exacerbated when instrumental noise is also present in
light curves.

@@

In addition to this, there are several other aspects of the GP method we hope
to test in future, listed as follows:.

>>>>

\author{%
   Ruth Angus\altaffilmark{\columbia},
   Timothy Morton\altaffilmark{\princeton},
   Suzanne Aigrain\altaffilmark{\oxford},
   Daniel Foreman-Mackey\altaffilmark{\uw,\sagan}
   \& Vinesh Rajpaul\altaffilmark{\oxford}
}
\altaffiltext{\columbia}{Simons Fellow, Department of Astronomy, Columbia
University, NY, NY, RuthAngus@gmail.com}
\altaffiltext{\princeton}{Department of Astrophysical Sciences, Princeton
University,  Princeton, NJ}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
UK}
\altaffiltext{\uw}{Department of Astronomy, University of
Washington, Seattle, WA}
\altaffiltext{\sagan}{Sagan Fellow}

\begin{abstract}
The light curves of spotted, rotating stars often vary in a non-sinusoidal and
quasi-periodic fashion---spots move on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing
    effective inference, such as highly degenerate or high-dimensional
    parameter spaces.
% \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian
    Process (GP) with a quasi-periodic covariance kernel function.
Using this highly flexible model, we are able to sample from the posterior
    Probability Density Function (PDF) of the periodic parameter,
    marginalising over the other kernel hyperparameters using a Markov Chain
    Monte Carlo approach.
To test the effectiveness of this method, we infer rotation periods from
\naigrain\ simulated stellar light curves, demonstrating that our results
    are more accurate than both a sine-fitting periodogram and an
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs it will enable hierarchical
    studies involving stellar rotation, particularly those involving
    population modelling, such as inferring stellar ages, obliquities in
    exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

The brightness of a spotted, rotating star often varies in a non-sinusoidal and
Quasi-Periodic (QP) manner, due to active regions on its surface which
rotate in and out of view.
Complicated surface spot patterns produce non-sinusoidal variations,
and the finite lifetimes of these active regions and
differential rotation on the stellar surface produce quasi-periodicity.
A strictly periodic sinusoid is therefore not necessarily a good model choice
for these time-series.
A physically realistic model of the stellar surface
would, ideally,  perfectly capture the complexity of shapes
within stellar light curves as well as the quasi-periodic nature, allowing for
extremely precise probabilistic period recovery when conditioned on the data.
However, such physical models require many free parameters in order to
accurately represent a stellar surface, and some of these parameters are
extremely degenerate \citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In addition to global stellar parameters such as inclination and rotation
period, each spot or active region should have (at minimum) a longitude,
latitude, size, temperature and lifetime.
Considering that stars may have hundreds of spots, the number of free
parameters in such a model quickly becomes unwieldy, especially to explore its
posterior Probability Density Function (PDF).
Simplified spot models, such as the one described in \citet{Lanza2014} where
only two spots are modelled, have produced successful results; however, such
relatively inflexible models sacrifice precision.

Standard methods to measure rotation periods include detecting
peaks in a Lomb-Scargle \citep{Lomb1976, Scargle1982} (LS) periodogram
\citep[e.g.][]{Reinhold2013}, Auto-Correlation Functions (ACFs),
\citep{Mcquillan2013} and wavelet transforms \citep{Garcia2014}.
The precisions of the LS periodogram and wavelet methods are limited by the
suitability of the model choice: a sinusoid for the LS periodogram,
and a choice of mother wavelet, assumed to desribe the data
over a range of transpositions \citep[see, \eg][]{Carter2010},
for the wavelet method.
In contrast, the ACF method is much better suited to signals that are
non-sinusoidal.
In fact, as long as the signal is approximately periodic the ACF will
display a peak at the rotation period, no matter its shape.
A drawback of the ACF method, however, is that it requires data to be
evenly-spaced\footnote{\citet{Edelson1988} describe a method for computing
ACFs for unevenly-spaced data.}, which is not exactly the case with \Kepler\
light curves (although in many cases it can be approximated as uniformly
sampled) and not nearly the case for most ground-based measurements, such
as the future Large Synoptic Survey Telescope (\LSST).
An ACF is also an operation performed on the data rather than a generative
model of the data, and so is not inherently probabilistic.
This means that the effects of the observational uncertainties cannot be
formally propagated to constraints on the rotation period.
% Many rotation periods in the literature have been inferred by measuring the
% position of the first peak in an ACF, however this approach can be dangerous.
% The exponential decay in correlation can shift the peak position short-wards
% of its true value, leading to an underestimate of the period.
% We return to this point in section \textsection \ref{sec:discussion}.

@@

GPs are commonly used in the machine learning community and increasingly
in other scientific fields such as biology, geophysics and cosmology.
More recently, GPs have been used in the stellar and exoplanet fields within
astronomy, to capture stellar variability or instrumental systematics
\citep[see \eg][]{Gibson2012, Haywood2014, Dawson2014, Barclay2015,
Haywood2015, Evans2015, Rajpaul2015, Rajpaul2016, Aigrain2016}.
They are useful in regression problems involving any stochastic process,
specifically when the probability distribution for the process is a
multi-variate Gaussian.
If the probability of obtaining a dataset is a Gaussian in $N$ dimensions,
where $N$ is the number of data points, a GP can describe that dataset.
An in-depth introduction to Gaussian processes is provided in
\citet{Rasmussen2005}.

GP models parameterise the covariance between data points by means of a
parametric kernel function that models the autocorrelation behaviour of the
signal and defines the form of the covariance matrix.
As a qualitative demonstration, we present the time-series in Figure
\ref{fig:GP_example}: the \kepler\ light curve of KIC \kepexample.
This is a relatively active star that rotates once every $\sim$
\kepexampleperiod\ days, with stochastic variability typical of \kepler\ FGK
stars.
Clearly, data points in this light curve are correlated.
Points close together in time are tightly correlated, and points more
widely separated are loosely correlated.
A GP models this variation in correlation with the separation between data
points; that is, it models the {\it covariance structure} rather than the data
directly.
This lends GPs their flexibility---they can model any time
series with a similar covariance structure.
In addition, a very simple function can usually capture the covariance
structure of a light curve, whereas modelling the time series itself
might require much more complexity.
Figure \ref{fig:GP_example} demonstrates how a GP model fits the light curve of
KIC \kepexample.
% Both provide adequate fits to the data, however only the periodic kernel
% function, `QP' is a useful model because it has a periodic parameter.
% I return to this point shortly.

A range of kernel functions can describe stellar variability.
For example, the simplest and most commonly used kernel function, the `Squared
Exponential' (SE), defined as follows, could adequately fit the
KIC \kepexample\ light curve:
\begin{equation}
k_{i,j} = A \exp \left(-\frac{(x_i - x_j)^2}{2l^2} \right).
\end{equation}
\label{eq:SE}
Here $A$ is the amplitude of covariance, $l$ is the length scale of
exponential decay, and $x_i-x_j$ is the separation between data points.
The SE kernel function has the advantage of being very simple, with
just two parameters, $A$ and $l$.
If $l$ is large, two data points far apart in $x$ will be tightly correlated,
and if small they will be loosely correlated.
Another property of the SE kernel function is that it produces functions that
are infinitely differentiable, making it possible to model a data set
and its derivatives simultaneously.
% \citep[see \eg][]{Rajpaul2015}
However, The SE kernel function does not well describe the covariance
in stellar light curves, nor is it {\it useful} for the problem of
rotation period inference because it does not capture periodic behaviour.
Inferring rotation periods thus requires a periodic kernel
function.
For this reason, we use the `Quasi-Periodic' kernel.
\citet{Rasmussen2005} model QP variability in CO$_2$ concentration on the
summit of the Mauna Loa volcano in Hawaii \citep[data from][]{Keeling2004}
using a kernel which is the product of a periodic and a SE kernel: the QP
kernel.
This kernel is defined as
\begin{equation}
k_{i,j} = A \exp \left[-\frac{(x_i - x_j)^2}{2l^2} -
\Gamma^2 \sin^2\left(\frac{2\pi}{P}\right) \right].
\end{equation}
\label{eq:QP}
It is the product of the SE kernel function, which describes the overall
covariance decay, and an exponentiated, squared, sinusoidal kernel function
that describes the periodic covariance structure.
$P$ can be interpreted as the rotation period of the star, and $\Gamma$
controls the amplitude of the $\sin^2$ term.
If $\Gamma$ is very small, only points almost exactly one period away are
tightly correlated and points that are slightly more or less than one period
away are very loosely correlated.
If $\Gamma$ is large, points separated by one period are tightly
correlated, and points separated by slightly more or less are still highly
correlated, although less so.
This kernel function allows two data points that are separated in time by one
rotation period to be tightly correlated, while also allowing
points separated by half a period to be weakly correlated.
We also use an extra parameter $\sigma$ which is an additional white noise
term added to the diagonal elements of the covariance matrix.
This can be interpreted to represent underestimation of observational
uncertainties~---~if the uncertainties reported on the data are too small, it
will be non-zero~---~or it can capture any remaining ``jitter'' or residuals
that are not captured by the effective GP model.
%In practice, this parameter should always be non-zero when performing
%GP regression, because the covariance matrix must be positive definite and
%slight deviations from positive definiteness can arise due to most
%solvers using approximate, rather than exact, matrix inversion.
%Including a small amount of extra variance in the model ensures enough
%flexibility for matrix inversion algorithms to not run into such numerical
%issues.
We use this QP kernel function to produce the GP model
that fits the \Kepler\ light curve in in Figure \ref{fig:GP_example}.

@@

This paper is laid out as follows.
In \textsection \ref{sec:simulations} we describe the simulated light curves
used to test the three rotation period recovery methods (ACF, LS periodogram
and GP).
The GP method is described in \textsection \ref{sec:method} and the ACF and LS
periodogram methods are described in \textsection \ref{sec:comparison}.
In \textsection \ref{sec:kepler} we apply our method to real \kepler\ data,
and the results are discussed in \textsection \ref{sec:discussion}.

@@

In order to test the GP method we attempted to measure the rotation periods of
a set of simulated light curves.
We used \naigrain\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \naigraincurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \naigrain\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in table \ref{tab:simulation_parameters}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
A histogram of the \naigrain\ solid-body rotation periods is shown in figure
\ref{fig:period_hist}.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real, quiet \kepler\ light
curves --- with no obvious astrophysical variability --- in order to preserve
the noise properties of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated light curve with a
period of \aigrainexampleperiod\ days.

@@

In order to recover the rotation period of a simulated light curve using
Gaussian processes, we sample the following posterior PDF:
\begin{equation}
p({\bm \theta}\,|\,y) \propto p(y\,|\,{\bm \theta}) p({\bm \theta}),
\end{equation}
\label{eq:posterior}
where $y$ are the light curve data, $\bm \theta$ are the hyperparameters
of the kernel described in equation \ref{eq:QP}, $p(y\,|\,{\bm \theta}) =
\mathcal{L}$, is the GP likelihood function, and $p({\bm \theta})$ is the prior
 on the hyperparameters.
The GP likelihood is similar to the simple Gaussian likelihood function used
for optimisation problems where the uncertainties are Gaussian and
uncorrelated.
The latter can be written
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\sum_{n=1}^N\frac{(y_n-\mu)^2}{\sigma_n^2}
    - \frac{N}{2}\ln(2\pi\sigma_n^2),
\end{equation}
\label{eq:chi2}
where $y_n$ are the data, $\mu$ is the mean model and $\sigma_n$ are the
Gaussian uncertainites on the data.
The equivalent equation in matrix notation is
\begin{equation}
\ln \mathcal{L} = -\frac{1}{2}\bf{r}^T\bf{C}^{-1}\bf{r}-\ln|\bf{C}|
    + \mathrm{constant},
\end{equation}
\label{eq:lhf1}
where $\bf{r}$ is the vector of residuals and $\bf{C}$ is the covariance
matrix,
\begin{eqnarray}
	\mathbf{C} &=& \left (\begin{array}{cccc}
	\sigma^2_1 & \sigma_{2, 1} & \cdots & \sigma_{N, 1} \\
	\sigma_{1, 2} & \sigma^2_2 & \cdots & \sigma_{N, 2} \\
    && \vdots & \\
	\sigma_{1, N} & \sigma_{2, N} & \cdots & \sigma^2_N
\end{array}\right )
\end{eqnarray}
In the case where the uncertainties are uncorrelated, the noise is `white',
(which is a frequent assumption made by astronomers and is sometimes
justified) and the off-diagonal elements of the covariance matrix are zero.
However, in the case where there is evidence for correlated
`noise'\footnote{In our case the `noise' is actually the model!  Incidentally, this approach is the reverse of the regression techniques
usually employed by astronomers.
In most problems in astronomy one tries to infer the parameters that describe
the mean model and, if correlated noise is present, to marginalise over that
noise.
Here, the parameters describing the correlated noise are what we are
interested in and our mean model is simply a straight line at $y=0$.}, as in the
case of \Kepler\ light curves, those off-diagonal elements are non-zero.
With GP regression, a covariance matrix generated by the kernel function,
${\bf K}$ replaces ${\bf C}$ in the above equation.

@@

Figures \ref{fig:compare_mcmc_acfprior} and \ref{fig:compare_mcmc_noprior}
summarize the results of this MCMC fitting procedure compared to the injected
`true' stellar rotation periods, for the uninformed and ACF-informed priors on
period, respectively.
In order to assess the performance of the GP and other period recovery
methods, we compute Median Absolute Deviations (MADs) of the results, relative
to the input periods.
We also compute the Median Relative Deviation (MRD), as a percentage.
These metrics are presented for the three different methods tested in this
paper in table \ref{tab:MADs}.
The informative and uninformative prior versions of the GP method have MRDs of
\percentgpMAD \% and \percentgpMADnp \% respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/25.png}
\caption{Marginal posterior PDFs of the QP GP model parameters, fit to the
    light curve in figure \ref{fig:demo_lc}.
    The green line in the period panel shows the injected period.
    This figure was made using \texttt{corner.py} \citep{Corner}.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

\section{Comparison with literature methods}
\label{sec:comparison}
\subsection{ACF}

@@

For each simulated light curve, a LS periodogram
\footnote{LS periodograms were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
was computed over a grid of 10,000 periods, evenly spaced in frequency,
between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
The MRD of the pgram-recovered periods is \percentpgramMAD\% (see table
\ref{tab:MADs} for a side-by-side comparison with the other methods).
There is a large number of drastically overestimated periods recovered using
the periodogram.
Excess power at long periods can be generated by long-term trends that are not
related to rotation.
This problem becomes exacerbated when instrumental noise is also present in
light curves.

@@

In addition to this, there are several other aspects of the GP method we are
continuing to develop, listed as follows:.
